---
output:
  html_document: default
  pdf_document: default
  md_document:
    variant: gfm
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.path="demo/", fig.align = 'center',
                      message=FALSE, warning=FALSE, cache=TRUE,
                      echo=TRUE, results='hold', fig.show='hold')
```
# Simulating fixation durations with the UCM model in R

## Introduction

This file demonstrates how to use the UCM model contained in the file
`UCM.R`. This implementation of the model was written with the
`simmer` package in `R`, which is a discrete event simulator. First,
let's load some packages, load the source code for the UCM utility
functions, and set a seed so that our results are reproducible:

```{r libraries}
library(simmer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(parallel)
library(viridis)

## Load the UCM source code
source('UCM.R')
set.seed(1234)
```

## Simulating a single trial using UCM
First, we can run UCM for five seconds with the default parameter settings:
```{r single_trial}
ucm <- UCM() %>%
    run(until=5)

ucm
```

As we can see, the current time of the simulator (now) is now 5
seconds, and the next event occurs at 5.01 seconds. This output also
tells us that the timer has cycled 21 times, 20 of which made it to
the labile stage, and 16 of which continued to saccade execution.

To get more detailed information, we can use the function
`get_states`, which returns a dataframe detailing the time of every
state transition in the model. The `stage` column tells us whether the
transition happened in the timer, the labile stage, the non-labile
stage, the motor planning stage, or saccade execution. The `n` column
tells you how many walks completed this stage at this point in time,
and the `id` column is a unique identifier linking random walks
between stages. The `state` column tells us which state the model
transitioned to at that point in time (from 0 to N-1), and `cum_state`
carries the same information except that each stage starts a state
above the previous stage, which is useful for plotting. Finally, the
`replication` column tells us which instance of the UCM this state
transition pertains to. Since we only ran one instance, `replication`
is always 1.

```{r get_states}
get_states(ucm)
```

To get information about the model's simulated fixations, we can use
the function `get_fixations`. The `id`, `n`, and `replication` columns
are the same as before. We also have columns telling us the
`start_time`, the `end_time`, and the `duration` of every fixation the
model has made.

```{r get_fixations}
get_fixations(ucm)
```

To get information about which saccade program cancellation, we can
use the function `get_cancellations`. The `id` and `replication`
columns are the same as before. The `cancelled` column tells us
whether the saccade program with this `id` was cancelled (1) or not
(0). Finally, `n_cancellations` tells us how many saccade programs
were cancelled during the fixation with this `id`. `n_cancellations`
is 0 for all cancelled saccade programs.

```{r get_cancellations}
get_cancellations(ucm)
```

We can also easily get a nifty plot of the UCM's state through time:

```{r trace_plot}
trace_plot(ucm)
ggsave('plots/UCM-trace.png', width=10, height=3)
```

## Custom parameter settings
We're also not limited to the default parameter settings. We can set
custom parameters like so:

```{r custom_parameters}
ucm <- UCM(N_timer=15, t_timer=.3,
           N_labile=20, t_labile=.2,
           N_nonlabile=20, t_nonlabile=.05,
           N_motor=30, t_motor=.02,
           N_execution=30, t_execution=.015) %>%
    run(until=5)

trace_plot(ucm)
ggsave('plots/UCM-trace2.png', width=10, height=3)
```


## Simulating many trials in parallel
Thankfully, `simmer` makes it easy to run multiple instances of the
`UCM` in parallel with the function `parallel::mcapply`. Depending on
how many cores you have on your machine, though, this could take a
minute or two. Also note that we need to use the `wrap` function,
which ensures that our simulation data is available after the parallel
processes are complete.

```{r parallel}
## set the number of cores
options(mc.cores=parallel::detectCores())

## run 1000 trials, each lasting 30s, in parallel
ucms <- mclapply(1:1000, function (i) {
    UCM() %>%
        run(until=30) %>%
        wrap()
})

head(ucms, n=3)
```

As you can see, `ucms` is a list containg 1000 instances of the UCM. We can use
the same functions as before on this list to, say, make a histogram of
fixation durations as a function of the number of cancellations within
the fixation:

```{r cancellation_hist}
cancellations_hist(ucms, max_cancellations=3)
ggsave('plots/UCM-cancellations.png', width=6, height=4)
```


## Local Rate Modulation
Since `UCM` is implemented with the `simmer` package, it is painless
to add more events to our simulator. One useful example is in local
rate modulation, where you want to speed up or slow down the random
walks of the timer, the labile stage, and the non-labile stage for
some period of time. To do so, we first need to define a `simmer`
trajectory which specifies when and how long to modulate the walk
rate. The UCM responds to the global attribute `"modulation"`, which
can be modified using the `simmer` command `set_global`. For instance,
the following code says that we want to reduce the rate by a factor of
4 for a duration of five seconds, then resume at the normal rate:

```{r modulation_trajectory}
modulation <- trajectory() %>%
    set_global("modulation", 0.25) %>%
    timeout(5) %>%
    set_global("modulation", 1)
```

Next, we can run the UCM as before, but this time we add a `simmer`
generator for our modulation trajectory, starting at five
seconds. Including the argument `mon=2` tells `simmer` to log the 
modulation events we defined above in case we need it for later.

```{r local_modulation}
ucm <- UCM() %>%
    add_generator('mod.', modulation, at(5), mon=2) %>%
    run(until=15)
```

To demonstrate how this alters the model's behavior, we can make a
trace plot as before, adding a shaded area to highlight the period of
modulation:

```{r modulation_plot}
trace_plot(ucm) +
    geom_rect(aes(xmin=5, xmax=10, ymin=-Inf, ymax=Inf, fill='modulation'),
              data=data.frame(), inherit.aes=FALSE, alpha=0.5, show.legend=FALSE)
ggsave('plots/UCM-local-modulation.png', width=10, height=3)
```


## Local Cancellation
The other major case where we need to augment the simulator with
additional information is when we want to model local cancellation of
saccade programs, for instance when a stimulus disappears for a
duration. To model this, we can create a trajectory which logs the
state of the stimulus using the global attribute `"stimulus"` (1 is on
and 0 is off). Then to make the appearance and disappearance of the
stimulus cancel any labile saccade programs, we can use the `simmer`
command `send('cancel-labile')`, which broadcasts a signal that
saccade programs in the labile stage listen to:

```{r stimulus}
stimulus <- trajectory() %>%
    set_global("stimulus", 0) %>% ## turn off the stimulus
    send('cancel-labile') %>%     ## cancel any labile programs
    timeout(1) %>%
    set_global("stimulus", 1) %>% ## turn on the stimulus
    send('cancel-labile')         ## cancel any labile programs
```

Then as before, we simply create an instance of the UCM, add a
generator to initialize the starting state of the stimulus to on, add
a generator to turn off the stimulus after two seconds, and run the model:

```{r local_cancellation}
ucm <- UCM() %>%
    add_generator('stim.', trajectory() %>% set_global('stimulus', 1), at(0)) %>%
    add_generator('stim-off.', stimulus, at(2), mon=2) %>%
    run(until=5)

trace_plot(ucm) +
    geom_rect(aes(xmin=2, xmax=3, ymin=-Inf, ymax=Inf, fill='modulation'),
              data=data.frame(), inherit.aes=FALSE, alpha=0.5, show.legend=FALSE)
ggsave('plots/UCM-local-cancellation.png', width=10, height=3)
```


## Conclusions
Hopefully this gives you a good idea of how to simulate fixation
durations using the UCM model. If you have any questions, feel free to
reach out at [kevin.oneill@duke.edu](mailto:kevin.oneill@duke.edu).
